---
title: "scRNA-seq QC"
output:
    html_document:
        code_folding: hide
        toc: true
        toc_float: true
        toc_depth: 3
        df_print: paged
---

```{r}
knitr::opts_chunk$set(
    warning=FALSE,
    message=FALSE,
    cache.lazy=FALSE)
```

```{r libraries}
library(Seurat)
library(BPCells)
library(scDblFinder)
library(tidyverse)
library(patchwork)
library(future)
library(parallel)
library(BiocParallel)
library(Matrix)
library(DT)
library(UpSetR)

source('../helpers.R')
```

```{r future_parallelization}

# ------------------------------------------------------------------------------
# Set up parallel computing
# ------------------------------------------------------------------------------

# Assign the number of cores for parallelization
ncpus <- future::availableCores()

# Adjust max memory allowed
options(future.globals.maxSize=500000 * 1024^2,   # 500G (1000 * 1024^2 = 1G)
        mc.cores=ncpus,
        MulticoreParam=quote(MulticoreParam(workers=ncpus)),
        future.rng.onMisuse="ignore")

# Parallelize the run
plan("multicore", workers=ncpus)

register(MulticoreParam(workers=ncpus))
```

# Overview

This document details the QC stages of scRNA-seq analysis and runs PCA and UMAP on 
both the integrated and non-integrated data after SCTransform normalization.

The final SeuratObject is saved at the end, which can be loaded into downstream 
clustering and marker gene identification analyses, which happen in other documents.
This lets us prepare the data just once here and then try out multiple parameters 
downstream independently.

```{r config, cache=TRUE}

# ------------------------------------------------------------------------------
# User configuration:
# - input/output files and directories
# - user variables
# ------------------------------------------------------------------------------

# Assign file path to sampletable
sampletable <- '../sampletable.tsv'
sample_paths <- get_sample_paths(sampletable)

# Assign factor levels
age.levels <- c(
    '6w',
    '10w',
    '12w',
    '14w',
    '18w')

genotype.levels <- c('WT', 'cKO')
sex.levels <- c('M', 'F')


# Assign metric names associated with scDblFinder
doublet_class <- 'scDblFinder.class'
doublet_score <- 'scDblFinder.score'

# Set regex patterns to detect mitochondrial/ribosomal genes. This may need to be
# tweaked based on the species being analyzed.
#
# mito_pattern matches 'mt-**', 'Mt_**', 'MT-**', 'mt:**'
# ribo_pattern matches 'Rps**', 'Rpl**', 'RPS**', 'RPL**', etc
mito_pattern <- '^mt-|Mt_|MT-|mt:'
ribo_pattern <- '^R[p|P][s|S]*[l|L]*'

# Set qc metrics to calculate
qc_metrics <- c('nFeature_RNA', 'nCount_RNA', 'percent.mito')

# Which qc metrics to plot
plot_metrics <- c('percent.mito', 'percent.ribo',
                  'nFeature_RNA', 'nCount_RNA')

# Set number of variable features for performing normalization
variable.features.n <- 3000

# Pick assay to use for integration
# - 'SCT': normalizes using `SCTransform`
# - 'RNA': normalizes using `LogNormzlize`
selected_assay <- 'SCT'

# Decide whether not to run SCTransform v2
sct.v2 <- TRUE

# Dimension reduction assays
# By default, this only contains the assay selected above
dimred_assays <- selected_assay

# Set max number of PCs to calculate. Used in RunPCA()
pca_dims <- 60

# Set number of UMAP dimensions to calculate. Used in RunUMAP()
umap_dims <- 40

pref <- 'seurat-v5-qc-sct-sketch'


# Create a subdirectory to store BPCells matices 
bpc.dir <- paste0(pref, '-BPCells')
if (!dir.exists(bpc.dir)) {
    # Create a new one if there's no pre-existing directory
    dir.create(bpc.dir)
} else {
    # Remove and create a new one if there's already pre-existing directory
    # NOTE: this is required due to errors raised by `wirte_matrix_dir`
    #       when finding a directory with the same name
    unlink(bpc.dir, recursive=TRUE)
    dir.create(bpc.dir)
}


# Determine whether you'll run integration on sketched data
# NOTE: Set to TRUE if your input has more than 2^31-1
sketch.dataset <- TRUE
```

# Read data

Data are imported from the output directories created by 10X Genomics’ Cell Ranger 
into a list of Seurat objects, one per sample.

```{r read_data, cache=TRUE, dependson='config'}

# ------------------------------------------------------------------------------
# Cellragner-generated `h5` files are imported and saved as BPCells 
# matrices on disk
# 
# See https://satijalab.org/seurat/articles/seurat5_bpcells_interaction_vignette#load-data-from-one-h5-file
# ------------------------------------------------------------------------------

# Read and create a list of seurat obj
obj.list <- mclapply(names(sample_paths), function(name) {

    # Create a seurat obj with in-memory dgCMatrix
    path <- sample_paths[name]
    sc.data <- Read10X(path)
    obj <- CreateSeuratObject(sc.data)

    # Write the matrix on-disk
    # (Overwrites pre-existing matrix!)
    ondisk.path <- file.path(bpc.dir, name)
    if (dir.exists(ondisk.path)) {
        unlink(ondisk.path, recursive=TRUE)
    }
    write_matrix_dir(mat=obj[['RNA']]$counts, dir=ondisk.path)

    # Replace the count matrix with on-disk `BPCells` matrix
    obj.mat <- open_matrix_dir(ondisk.path)
    obj[['RNA']]$counts <- obj.mat

    # Update the `orig.ident` column in the metadata to samplename
    obj$orig.ident <- name

    # Add metadata columns and convert each column to `factor if necessary
    obj@meta.data <- obj@meta.data %>%
        separate(orig.ident,
            c('age', 'genotype', 'sex', 'rep'),
            remove=FALSE) %>%
        mutate(genotype=factor(genotype, levels=genotype.levels),
            sex=factor(sex, levels=sex.levels),
            age=factor(age, levels=age.levels),
            orig.ident=factor(orig.ident, levels=names(sample_paths))
        )
    return(obj)
    }
) %>%
    set_names(names(sample_paths))

```

# QC

This section shows the various calculated QC metrics and thresholds.


## Cliff-knee plot {.tabset}

Here we show a plot of UMI counts vs barcodes, also known as the `cliff-knee`
plot, for each sample.

- The color of the line indicates whether a particular barcode
  was classified by `cellranger` as coming from an actual cell
  (`black`) or was part of the background (`gray`).
- The cells colored in `black` form the starting point of our
  analysis workflow.

```{r cliff_knee, results='asis', cache=TRUE, dependson='read_data'}

# ------------------------------------------------------------------------------
# Create Cliss-knee plots
# ------------------------------------------------------------------------------

p_list <- mclapply(sample_paths, get_cliff_knee) %>%
    set_names(names(obj.list))

for(s in names(p_list)){
    mdcat(s, level=3)
    print(p_list[[s]])
    cat('\n\n')
}

```

## Doublet detection

Doublets are detected using the `scDblFinder` package. This finds possible
doublets missed by Cell Ranger's cliff-knee plots; these will be filtered out.

```{r doubletfinding, dependson='read_data', cache=TRUE}

# -------------------------------------------------------------------------------
# Use scDblFinder package to identify doublets. This is done on the raw counts,
# and for each sample separately. We need to temporarily convert to SCE for the
# purposes of doublet detection. The resulting SeuratObj has additional columns:
#
# - scDblFinder.class
# - scDblFinder.score
# - scDblFinder.weighted
# - scDblFinder.cxds_score
# ------------------------------------------------------------------------------


# NOTE:
# - The `BPCells` matrix is not compatible with `sce`.
# - We conver the count layer with an on-disk `BPCells` matrix to an in-memory `dgCMatrix`.
obj.list <- mclapply(names(obj.list), function(name) {

    # Converts on-disk BPCells matrix to in-memory dgCMatrix in each seurat obj
    x <- obj.list[[name]]
    counts <- as(object=x[['RNA']]$counts, Class='dgCMatrix')
    x[['RNA']]$counts <- counts

    # Convert seurat obj to sce obj
    sce <- as.SingleCellExperiment(x)

    # Compute doublets
    sce <- scDblFinder(sce)

    # Slice columns created by scDblFinder in the `colData`
    dbl.df <- colData(sce) %>%
        as.data.frame() %>%
        dplyr::select(starts_with('scDblFinder'))

    # Add the new columns to the seurat j
    x@meta.data <- cbind(x@meta.data, dbl.df)

    # Convert in-memory `dgCMatrix` to on-disk `BPCells` matrix
    ondisk.path <- file.path(bpc.dir, name)
    obj.mat <- open_matrix_dir(ondisk.path)
    x[['RNA']]$counts <- obj.mat

    return(x)
    }
) %>%
    set_names(names(obj.list))
```

```{r print_doublets}
data.frame(
    count=sapply(obj.list, function(x) sum(x$scDblFinder.class=='doublet')),
    total=sapply(obj.list, function(x) ncol(x))
) %>%
    mutate(percent=round(count/total*100, 2)) %>%
    dplyr::select(-total) %>%
    knitr::kable()
```


## Quality metrics {.tabset}

Here we use the `scuttle::isOutlier()` function on various metrics. This
automatically selects a threshold, flagging cells that exceed 3 median absolute
deviations. This avoids qualitative, *ad hoc* filtering choices, and lets use
thresholds in a dataset-specific manner.

Note that in cases where the lower value is negative and the metrics only take
positive values, there will not be a negative threshold.

The table shows the values; the plot shows horizontal lines for each threshold
for each sample. Only metrics listed in the table (and those that have
horizontal lines) are used for filtering purposes.

```{r thresholds, results='asis', dependson='doubletfinding', cache=TRUE}
# ------------------------------------------------------------------------------
# - Calculate the percentage of mitochondrial and ribosomal gene expression
# - Update metadata with columns indicating outlier cells to be removed
# ------------------------------------------------------------------------------

obj.list <- lapply(obj.list,
    function(x) add_percentage_features(
        x,
        mito_pattern=mito_pattern,
        ribo_pattern=ribo_pattern,
        mito_label='percent.mito',
        ribo_label='percent.ribo')
)

# TODO: add check to see if all percent.mito/percent.ribo == 0

# Use isOutlier() to detect cells with MAD > 3 and store in data.frame `m`.

# NOTE: recall that in purrr parlance, _dfr suffix means "glue together the
# returned dataframes, row-wise". imap provides both the object and the name
# (as opposed to just the object for map) which allows us to properly add the
# orig.id column.
m <- do.call('rbind',
    lapply(names(obj.list),
           function(x) find_thresholds(obj.list[[x]], x, metrics=qc_metrics)
    )
)

metrics_df <- purrr::map_dfr(obj.list, function (x) x@meta.data) %>%
  dplyr::select(all_of(plot_metrics), orig.ident) %>%
  reshape2::melt()


m2 <- m %>%
    rename(orig.ident=orig.id, variable=metric)

# Apply thresholds based on `m`
for (ident in unique(m$orig.id)){
   mdata <- obj.list[[ident]]@meta.data

   mdata$any_outlier <- FALSE
   for (metric.i in unique(m$metric)){
     s <- dplyr::filter(m, orig.id==ident, metric==metric.i)
     above <- mdata[, metric.i] > s$higher
     mdata[, paste0(metric.i, '_outlier')] <- above
     mdata$any_outlier <- mdata$any_outlier | above
   }

   # We also want to remove doublets.
   mdata$keep <- !(mdata$any_outlier | mdata[, doublet_class ] == 'doublet')

   obj.list[[ident]]@meta.data <- mdata
}
```

```{r display_qc_thres, results='asis'}

# ------------------------------------------------------------------------------
# Clean and display QC results
# ------------------------------------------------------------------------------

# reformat table of QC metric thresholds
m_new <- list()
for(i in 1:nrow(m)){
    s <- m[i, 'orig.id']
    tmp <- setNames(m[i, c('lower', 'higher')],
                    paste0(m[i, 'metric'], '_', c('lower','higher')))
    if(!s %in% names(m_new)){
      m_new[[s]] <- tmp
    } else {
      m_new[[s]] <- c(m_new[[s]], tmp)
    }
}

m_new_df <- as.data.frame(do.call('rbind', m_new))
m_new_df <- cbind(sample=rownames(m_new_df), m_new_df)

# build sketch to show thresholds
sketch <- htmltools::withTags(table(
    class = 'display',
    thead(
        tr(
            th(rowspan=2, 'sample'),
            lapply(qc_metrics,
                   function(x) th(class='dt-center', colspan=2, x))
        ),
        tr(
            lapply(rep(c('lower','higher'), length(qc_metrics)), th)
        )
    )
))

# display table using sketch

# NOTE:
#
# - columnDefs uses 0-based column indexing
# - formatSignif uses 1-based column indexing
mdcat('Table', level=3)
m_new_df %>%
    datatable(rownames=FALSE,
        container=sketch,
        options=list(
            columnDefs=list(list(className='dt-center',
                                 targets=1:(ncol(m_new_df)-1))))
        ) %>%
    formatSignif(columns=2:ncol(m_new_df),
                 digits=4)
cat('\n\n')
```


```{r qc_metric_plot, results='asis', fig.width=10, fig.height=8}

# ------------------------------------------------------------------------------
# Plot QC results
# ------------------------------------------------------------------------------

# plot metric thresholds
metrics_df <- metrics_df %>%
    mutate(orig.ident=factor(orig.ident, levels=names(obj.list)))
mdcat('Plot {.tabset}', level=3)
for (var in unique(metrics_df$variable)) {
    mdcat(var, level=4)
    p <- ggplot(metrics_df %>% dplyr::filter(variable == var)) +
      aes(y=value, x=orig.ident, fill=orig.ident) +
      theme(axis.text.x=element_text(angle=90, vjust=0.1, hjust=1)) +
      geom_violin() +
      geom_hline(data=m2[m2$variable == var,],
                 aes(yintercept=higher, color=orig.ident))

    print(p)
}


```

## UpSet plots of filtered cells {.tabset}

Here we show the distribution of cells that were filtered out due to possibly
multiple metrics.

```{r upset_plot, results='asis'}

# ------------------------------------------------------------------------------
# Plot outlier cells being unique and intersecting across the QC metrics
# ------------------------------------------------------------------------------

# UpSet plots of filtered cells
filtered_matrix <- list()
for (name in names(obj.list)){
    mdcat(name, '{.tabset}', level=3)
    metric_cols <- sapply(qc_metrics, function(x) paste0(x, '_outlier'))

    mdata <- obj.list[[name]]@meta.data
    filtered_matrix[[name]] <- cbind(
        mdata[, metric_cols],
        doublet=mdata[, doublet_class] == 'doublet')

    print(upset(filtered_matrix[[name]] + 0,
                text.scale=1.5))
}
```

## Filtering summary

Here we summarize the number of cells that were filtered out
based on quality metrics and doublet detection.

```{r filtering_summary}
# ------------------------------------------------------------------------------
# Print QC summary table
# ------------------------------------------------------------------------------

filt <- lapply(names(obj.list), function(x){
    mdata <- obj.list[[x]]@meta.data
    raw <- nrow(mdata)
    ff <- colSums(filtered_matrix[[x]])
    filtered <- sum(mdata[, 'keep'])
    c(raw=raw,
      ff,
      filtered=filtered,
      filtered.pct=(raw - filtered)/raw)
})
df <- do.call('rbind', filt)
df <- cbind(sample=names(obj.list), df)
knitr::kable(df)

```

```{r save_qc}
qc_list <- list(
    metadata=lapply(obj.list, function(x) x@meta.data),
    thresholds=m %>% rename(orig.ident=orig.id))
saveRDS(qc_list, paste0(pref, '-qc.Rds'))
```

# Normalization

We perform normalization using either `SCTransform` (aka sctransformation) or 
`NormalizeData` implementing `LogNormalize` in Seurat v4 (aka log-normalization). 
Normalization is performed on each dataset independently.

The `SCTransform` function runs in two different versions:

- [v1](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-019-1874-1): v1 
normalizes raw unique molecular identifier (UMI) counts using negative binomial (NB) 
generalized linear models (GLM) for each gene as the response variable and sequencing depth
as the explanatory variable in each dataset. Parameters for the GLM is further learned by 
accounting for a gene's average expression, which is called _regularized NB regression_.
- [v2](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-021-02584-9): v2 
normalizes the counts using the same algorithm as the v1 except that the regularization of
parameters is performed after excluding genes with very low expression or where variance of
their molecular counts does not exceed the mean. Such genes violate assumptions of NB distribution
and better fit Poisson distribution.

In log-normalization, we normalize single cells by dividing feature counts per cell by the total 
counts for that cell, followed by multiplication with the `scale.factor`, and then transforming 
to the natural log scale using `log1p`.


```{r normalization_option}
norm_function <- ifelse(
    selected_assay == 'SCT',
    'SCT',
    'LogNormalize'
)

print(paste0('Normalization using ', norm_function, ' performed...'))
if (norm_function == 'SCT') {
    print(
        paste0("Calling sctransform version ", ifelse(sct.v2, "v2", "v1"), "...")
        )
}
```


```{r prep_normalization, dependson='thresholds', cache=TRUE}
# ------------------------------------------------------------------------------
# - Remove outlier cells
# - Merge datasets
# ------------------------------------------------------------------------------

# Remove outlier cells
obj.list <- mclapply(obj.list, function(x) subset(x, subset=keep))

if (length(obj.list) > 1) {

    # Merge all datasets
    obj.merged <- merge(
        x=obj.list[[1]],
        y=obj.list[2:length(obj.list)],
    )
    sct.input <- obj.merged
} else {
    sct.input <- obj.list[[1]]
}
```


```{r normalize_data, cache=TRUE, dependson='prep_normalization'}

# ------------------------------------------------------------------------------
# - Normalize counts using `SCTransform`
#     - If set to 'v2', it causes the model to learn theta and intercept only besides
#       excluding poisson genes from learning and regularization
#     - SCTransform() replaces NormalizeData(), ScaleData(), and FindVariableFeatures()
#     - https://satijalab.org/seurat/articles/sctransform_v2_vignette
#     - https://satijalab.org/seurat/reference/sctransform
# - Normalize counts using `NormalizeData` with the `normalization.method` arg 
#   set to 'LogNormalize'
# ------------------------------------------------------------------------------

if (selected_assay == 'SCT') {

    # Normalize and find variable features
    # NOTE: When the `vst.flavor` is set to 'v2', following arguments are
    # automatically adjusted:
    # - method = glmGamPoi_offset
    # - n_cells = 2000,
    # - exclude_poisson = TRUE 
    # It causes the model to learn theta and intercept only besides excluding 
    # poisson genes from learning and regularization
    # NOTE: When having the object split into multiple layers, `SCTransform` 
    # runs on each count layer separately
    obj.merged <- SCTransform(
        sct.input,
        vst.flavor=ifelse(sct.v2, 'v2', NULL),
        verbose=FALSE)

} else if (selected_assay == 'RNA') {

    # Normalize and find variable features
    obj.merged <- NormalizeData(
        obj.merged,
        normalization.method="LogNormalize",
        verbose=FALSE) %>%
        # Retrieve N variable features
        FindVariableFeatures(
            nfeatures=variable.features.n,
            verbose=FALSE) %>%
        ScaleData(verbose=FALSE)
} else {
    stop('Invalid normalization method!')
}


# Retrieve N variable features computed by `SCTransform()`
features <- VariableFeatures(obj.merged)

# Reduce dimensions
obj.merged <- RunPCA(
    obj.merged,
#     assay=selected_assay,
#     features=features,
    reduction.name=paste0(tolower(selected_assay), '.pca'),
    reduction.key=paste0(tolower(selected_assay), 'pca_'),
    verbose=FALSE,
    npcs=pca_dims) %>%
    RunUMAP(
#        assay=selected_assay,
       reduction=paste0(tolower(selected_assay), '.pca'),
       reduction.name=paste0(tolower(selected_assay), '.umap'),
       reduction.key=paste0(tolower(selected_assay), 'umap_'),
       verbose=FALSE,
       dims=1:umap_dims)
```


# Sketching representative cells (optional)

The [sketching](https://www.sciencedirect.com/science/article/pii/S2405471219301528) method 
enables the integration of large-scale datasets with over 2^31-1 (2,147,483,647) non-zero entries.
It samples a subset of representative cells using a _leverage score_, which reflects the magnitude
of its contribution to the gene-covariance matrix. The _leverage score_ is highest for rare
populations in a dataset. Therefore, the sketched analysis will oversample rare populations,
thereby retaining the biological complexity of the sample while drastically compressing the dataset.


```{r sketch_option}
# Print
print(paste("Sketch performed:", sketch.dataset))
```

```{r sketch, dependson='normalize_data', cache=TRUE}
# ------------------------------------------------------------------------------
# - Subset representative cells using `SketchData`
# - https://satijalab.org/seurat/articles/seurat5_sketch_analysis
# - https://satijalab.org/seurat/articles/parsebio_sketch_integration
# - https://satijalab.org/seurat/reference/sketchdata
# ------------------------------------------------------------------------------

if (sketch.dataset) {

    if (length(obj.list) > 1) {
        if (selected_assay != 'RNA') {
            # Preprocess the obj
            # NOTE: `SketchData` only accepts log-normalized obj
            DefaultAssay(obj.merged) <- 'RNA'
            obj.merged <- NormalizeData(
                obj.merged,
                normalization.method='LogNormalize',
                verbose=FALSE) %>%
                FindVariableFeatures(
                    verbose=FALSE,
                    nfeatures=variable.features.n
                    )
        }

        # Run sketch
        obj.merged <- SketchData(
            obj.merged,
            verbose=FALSE,
            sketched.assay='sketch'
        )
    } else {
        print("You have a single dataset. Sketching is skipped.")
    }
}

```


# Integration

Integration is performed either with or without sketching. Without sketching, data is
integrated on total cells normalized using `SCTransform` or `LogNormalize`. With sketching,
integration is performed on a subset of representative cells (called sketched cells) across
all datasets in a low-dimensional shared space, and the integrated data is then propagated
back to the full dataset.
We employ [Canonical Correlation Analysis (CCA)](https://www.cell.com/cell/fulltext/S0092-8674(19)30559-8)
as the default method for integration.

After integration, sample-specific celltypes should remain separate in the
resulting UMAP, but there is always a risk of "regressing away" interesting
biology. So here we run both ways so that we can make an informed decision
about what to use.

The two versions are split out into two tabs in the below sections. The
non-integrated version uses the normalized counts computed using either `SCTransform`
or `LogNormalize` and is labeled "SCT" (for `SCTransform`) or "RNA" (for `LogNormalize`). For
these tabs, each sample was independently normalized, the data were combined
back together into a large matrix, and the PCA and UMAPs use that combined
matrix. In contrast, the "integrated" tabs have PCA and UMAP performed on
the back-propagated data.

Note that in all cases, differential expression and marker genes should be
identified using the `SCT` or `RNA` slot, not the `integrated` slot.

We run integration only if we have more than one dataset.


```{r print_integration_option}
print(paste0(
    "Integration performed on ",
    ifelse(sketch.dataset, "sketched", "normalized unsketched"),
    " data ..."))
```

```{r integration, cache=TRUE, dependson=c('sketch', 'normalize_data')}

# ------------------------------------------------------------------------------
# Integration is performed in two different cases below:
# - Integration of sketched data
# - Integration of sctransformed data
# - References:
#     - https://satijalab.org/seurat/articles/parsebio_sketch_integration
#     - https://satijalab.org/seurat/articles/seurat5_integration
#     - https://satijalab.org/seurat/articles/integration_introduction
# -----------------------------------------------------------------------------
# Run integration
if (length(obj.list) > 1) {

    # add 'integrated' assay to dimension reduction assays
    dimred_assays <- c(dimred_assays, 'integrated')

    # Integrate SCT (unsketched) data
    if (!sketch.dataset) {

        # Integrate datasets
        obj.integrated <- IntegrateLayers(
            obj.merged,
            method=CCAIntegration,
            orig.reduction=paste0(tolower(selected_assay), '.pca'),
            assay=selected_assay,
            new.reduction='integrated.cca',
            normalization.method=norm_function,
            verbose=FALSE
            )

    # Integrate sketched data
    } else {
        # Preprocess the sketched obj
        DefaultAssay(obj.merged) <- 'sketch'

        # Retrieve N variable features computed by `SCTransform()`
        obj.merged <- FindVariableFeatures(
            obj.merged,
            verbose=FALSE,
            nfeatures=variable.features.n)
        features <- VariableFeatures(obj.merged)

        # Scale and reduce dimensions
        obj.merged <- ScaleData(
            obj.merged,
            features=features,
            verbose=FALSE) %>%
            RunPCA(
            assay='sketch',
            features=features,
            reduction.name='sketch.pca',
            reduction.key='sketchpca_',
            verbose=FALSE) 

        # Integrate layers
        obj.integrated <- IntegrateLayers(
            obj.merged,
            method=CCAIntegration,
            orig.reduction='sketch.pca',
            assay="sketch",
            features=features,
            new.reduction='integrated.cca',
            verbose=FALSE
            )

        # Integrate embeddings from the sketched assay
        obj.integrated <- ProjectIntegration(
            obj.integrated,
            verbose=FALSE,
            sketched.assay='sketch',
            method='sketch',
            assay="RNA",
            reduction='integrated.cca') %>%
        # Project high dimensional scRNA expression data from a full dataset
        # onto the lower dimensional embedding of the sketch of the dataset
            ProjectData(
                sketched.assay='sketch',
                assay='RNA',
                normalization.method=norm_function,
                sketched.reduction='integrated.cca.full',
                full.reduction='integrated.cca.full',
                verbose=FALSE,
                dims=1:30
            )

    }

} else {
    print("There is only one dataset. It cannot be integrated.")
    obj.integrated <- obj.list[[1]]
    # remove 'integrated' from assays
    dimred_assays <- setdiff(dimred_assays, 'integrated')  
}

```


```{r integrated_dimred, cache=TRUE, dependson='integration'}
# ------------------------------------------------------------------------------
# Reduce dimensions on integrated data
# ------------------------------------------------------------------------------

integrated.reduction <- ifelse(
    sketch.dataset,
    'integrated.cca.full',
    'integrated.cca')

obj.integrated <- RunUMAP(
    obj.integrated,
    reduction=integrated.reduction,
    dims=1:umap_dims,
    reduction.name='integrated.umap',
    reduction.key='integratedumap_',
    verbose=FALSE)

obj.integrated <- RunPCA(
    obj.integrated,
    reduction=integrated.reduction,
    dims=1:pcs_dims,
    reduction.name='integrated.pca',
    reduction.key='integratedpca_',
    verbose=FALSE
    )

```

```{r save_integrated_obj}
saveRDS(obj.integrated, file=paste0(pref, '-integrated.rds'), compress=FALSE)
```


## Dimension loadings per gene {.tabset}

The following graphs show the genes that drove the most variance within the first 6 Principal
Component axes.

```{r dimension_loadings, results='asis', cache=TRUE, dependson='integrated_dimred'}
for(selected_assay in tolower(dimred_assays)){
    mdcat(selected_assay, ' {.tabset}', level=3)
    lst <- list('PCs 1 and 2'=1:2, 'PCs 3 and 4'=3:4, 'PCs 5 and 6'=5:6)
    for (name in names(lst)){
        mdcat(name, level=4)
        print(
            VizDimLoadings(object=obj.integrated, dims=lst[[name]],
                           reduction=paste0(selected_assay, '.pca'))
        )
    }
}

```

## Distribution of variance in principal components {.tabset}

A total of 60 dimensions were computed, but only the most statistically
significant ones should be included to reduce computational burden on the UMAP
and on any subsequent clustering. The threshold is determined visually from an
Elbow plot. Here, we chose the first `r pca_dims` at most.

```{r elbowplot, results='asis',  cache=TRUE, dependson='integrated_dimred'}

for(selected_assay in tolower(dimred_assays)){
    mdcat(selected_assay, ' {.tabset}', level=3)
    reduction <- paste0(selected_assay, ".pca")
    n.pcs <- length(obj.integrated[[reduction]]@stdev)
    ndims <- min(n.pcs, pca_dims)
    print(
        ElbowPlot(
            object=obj.integrated,
            ndims=ndims,
            reduction=reduction
        )
    )
}

```

## UMAP plots {.tabset}

Uniform Manifold Approximation and Projection (UMAP) is a manifold learning
technique for dimension reduction that is run on the Principal Component
Analysis (PCA). This is a way to look at sample clustering. The x- and y-axes
do not have units, rather, they represent the dimensions along which the
samples vary the most. The first two dimensions are plotted in the graph below.

Note that these data are not yet clustered; these integrated and non-integrated
versions are for assessing whether or not integration was useful for this
study.

```{r umap, results='asis', cache=TRUE, dependson='integrated_dimred', fig.width=10, fig.height=8}

for (prefix in tolower(dimred_assays)){
    mdcat(prefix, '{.tabset}', level=3)
    mdcat('Overlay', level=4)
    print(DimPlot(object=obj.integrated,
                  reduction=paste0(prefix, ".umap"),
                  group.by="orig.ident"))
    mdcat('Side-by-side', level=4)
    print(DimPlot(object=obj.integrated,
                  reduction=paste0(prefix, ".umap"),
                  split.by="orig.ident"))
}

```


## Feature plots {.tabset}

These UMAP plots show the various metrics calculated in the QC steps above
on the post-filtered dataset.

```{r feature_plots, results='asis', fig.height=16, fig.width=12, cache=TRUE, dependson='integrated_dimred'}

for (prefix in tolower(dimred_assays)){
    mdcat(prefix, level=3)
    print(FeaturePlot(
        obj.integrated,
        reduction=paste0(prefix, '.umap'),
        split.by='orig.ident',
        features=c(plot_metrics, doublet_score)
    ))
}

```

# Session info

```{r session_info, collapse=FALSE}
sessionInfo()
```

